# Data transformation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
```

## Crime Data

### Data Import - 2021 Data

We begin by importing our data source as-is. This will give us a starting point for our data adjustment and cleaning. We save this data as 'nyc_arrests'.

```{r}
nyc_arrests <- read.table("https://data.cityofnewyork.us/api/views/uip8-fykc/rows.csv?accessType=DOWNLOAD",
                          header=TRUE,
                          sep=",",
                          quote='"'
                          )
```

For easier use of the dates associated with the arrests, we will convert these into R-compatible dates. We will also remove several columns not necessary for the analysis.

```{r}
nyc_arrests$ARREST_DATE <- as_date(nyc_arrests$ARREST_DATE, format="%m/%d/%Y")
nyc_arrests <- nyc_arrests %>% select(-c(New.Georeferenced.Column, X_COORD_CD, Y_COORD_CD))
```

### Data Import - 2006-2021 Data

We now repeat this transformation on our older data. We save this data as 'nyc_arrests_old'. One issue with this dataset is its size. Weighing in at over one Gigabyte, this dataset in its unmodified form takes a very long time to download and modify. As a result, we have decided to utilize a reduced version of this dataset. The transformations used to create this historical dataset are seen below:

```{r, eval=FALSE}
nyc_arrests <- read.table("D:/Documents/R Projects/Final Project Proposal/NYPD_Arrests_Data__Historic_.csv",
                           header=TRUE,
                           sep=",",
                           quote='"'
)

nyc_arrests <- nyc_arrests %>% select(-c(Lon_Lat, X_COORD_CD, Y_COORD_CD))

nyc_arrests$ARREST_DATE <- as_date(nyc_arrests$ARREST_DATE, format="%m/%d/%Y")

nyc_arrests <- filter(nyc_arrests, ARREST_DATE > "2010-12-31")

write.csv(nyc_arrests, file="nyc_arrests_post2010.csv")
```

We then run a matching set of data transformations.

```{r}
nyc_arrests_old <- read.table("nyc_arrests_post2010.csv",
                          header=TRUE,
                          sep=",",
                          quote='"'
                          )
nyc_arrests_old <- nyc_arrests_old %>% select(-c(X))
nyc_arrests_old$ARREST_DATE <- as_date(nyc_arrests_old$ARREST_DATE, format="%Y-%m-%d")
```

### Combining Old and New

If we need a combined dataset, we can append the historical data to the 2021 data like so:

```{r}
nyc_arrests_combined <- bind_rows(nyc_arrests, nyc_arrests_old)
```

### Aggregating by Category

For many of our analyses, it is sensible to aggregate crime by its rough category, like so:

```{r}
nyc_arrests_by_category_year_boro <- (
  nyc_arrests_combined %>%
    drop_na(OFNS_DESC) %>%
    filter(OFNS_DESC != "", ARREST_BORO != "") %>%
    mutate(ARREST_YEAR = year(ARREST_DATE)) %>%
    group_by(ARREST_YEAR, OFNS_DESC, ARREST_BORO) %>%
    summarize(COUNT_OF_ARRESTS = n())
)
```



