# Data transformation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
```

## Data Import - 2021 Data

We begin by importing our data source as-is. This will give us a starting point for our data adjustment and cleaning. We save this data as 'nyc_arrests'.

```{r}
nyc_arrests <- read.table("https://data.cityofnewyork.us/api/views/uip8-fykc/rows.csv?accessType=DOWNLOAD",
                          header=TRUE,
                          sep=",",
                          quote='"'
                          )
```

For easier use of the dates associated with the arrests, we will convert these into R-compatible dates.

```{r}
nyc_arrests$ARREST_DATE <- as_date(nyc_arrests$ARREST_DATE, format="%m/%d/%Y")
```

## Data Import - 2006-2021 Data

We now repeat this transformation on our older data. We save this data as 'nyc_arrests_old'. One issue with this dataset is its size. Weighing in at over one Gigabyte, this dataset in its unmodified form takes a very long time to download and modify.

